# Customising plots {#sec-customising}

```{r setup-customising, include=FALSE}
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(janitor)
library(patchwork)
source("R/load_data.R")

port_summary <- dat_filter |>
  group_by(start_port) |>
  summarise(embarked = sum(total_embarked, na.rm = TRUE),
            .groups = "drop")
```

## Intended learning outcomes

By the end of this chapter you will be able to:

* Use `scale_*` functions to control axis labels, breaks and limits
* Distinguish `coord_cartesian()` from `scale_*` limits
* Apply built-in themes and set a global default with `theme_set()`
* Add titles, subtitles and captions with `labs()`
* Use the colour-blind-friendly `viridis` palettes

## Functions used

* ggplot2: `scale_x_discrete()`, `scale_x_continuous()`, `scale_y_continuous()`, `coord_cartesian()`, `scale_fill_viridis_d()`, `scale_colour_viridis_d()`, `labs()`, `theme_minimal()`, `theme_bw()`, `theme_classic()`, `theme_set()`
* scales: `label_comma()` (replacement for `scales::comma`)

## Scientific notation and number formatting

The column chart from @sec-basic-plots uses scientific notation on the y-axis because the numbers are so large.

```{r}
#| label: fig-col-scientific
#| fig-cap: "Total embarked, with the default y-axis in scientific notation."
#| fig-alt: "Column chart with Liverpool, London and Bristol on the x-axis and total embarked on the y-axis labelled in scientific notation (5e+05, 1e+06)."
ggplot(port_summary, aes(x = start_port, y = embarked, fill = start_port)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "E")
```

We can override that with `scale_y_continuous()` and a label-formatting function from the `scales` package.

```{r}
#| label: fig-col-comma
#| fig-cap: "Total embarked, with the y-axis formatted with thousands separators."
#| fig-alt: "Column chart with Liverpool, London and Bristol on the x-axis and total embarked on the y-axis labelled 0, 500,000, 1,000,000."
ggplot(port_summary, aes(x = start_port, y = embarked, fill = start_port)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "E") +
  scale_y_continuous(labels = scales::label_comma())
```

::: {.callout-note}
`scales::label_comma()` is the modern factory function. The older `scales::comma` (no parentheses) still works but is being phased out.
:::

## Axis labels and titles

`labs()` is the most concise way to set axis labels, the plot title, subtitle and caption.

```{r}
#| label: fig-col-labs
#| fig-cap: "Adding axis labels and a title."
#| fig-alt: "Column chart of total embarked by port with axis label 'Port of origin', y-axis label 'Number embarked', and title 'Number embarked on ships originating in British ports'."
ggplot(port_summary, aes(x = start_port, y = embarked, fill = start_port)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "E") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = "Port of origin",
       y = "Number embarked",
       title = "Number embarked on ships originating in British ports")
```

## Themes

`ggplot2` has several built-in themes. We have already set `theme_minimal()` as the global default in the project setup (see `R/my_setup.R`). You can override it for an individual plot by adding a different theme as a layer.

```{r}
#| label: fig-themes
#| fig-cap: "The same plot in three different themes."
#| fig-alt: "Three column charts of total embarked by port, each with a different theme applied: theme_minimal, theme_bw and theme_classic."
p <- ggplot(port_summary, aes(x = start_port, y = embarked, fill = start_port)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "E") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = NULL, y = NULL)

(p + theme_minimal() + ggtitle("theme_minimal")) +
  (p + theme_bw() + ggtitle("theme_bw")) +
  (p + theme_classic() + ggtitle("theme_classic"))
```

::: {.callout-tip}
Type `theme_` in the RStudio source pane and let auto-complete show you the available options. To set a theme for every subsequent plot in your script, run `theme_set(theme_bw())` near the top.
:::

## Discrete x-axis labels

`scale_x_discrete()` lets you control the tick labels on a categorical axis without renaming the underlying data.

```{r}
#| label: fig-discrete-labels
#| fig-cap: "Renaming x-axis ticks with scale_x_discrete()."
#| fig-alt: "Column chart with x-axis ticks renamed to 'Liverpool', 'Port of London' and 'Bristol'."
ggplot(port_summary, aes(x = start_port, y = embarked, fill = start_port)) +
  geom_col(show.legend = FALSE) +
  scale_x_discrete(labels = c("Liverpool", "Port of London", "Bristol")) +
  scale_fill_viridis_d(option = "E") +
  scale_y_continuous(labels = scales::label_comma())
```

## Axis limits the right way

There are two ways to restrict the visible range of an axis, and they behave differently.

`coord_cartesian(xlim = c(...), ylim = c(...))` zooms in on the plot. Data outside the range are still used in calculations (such as drawing a smoother) and are simply cropped from view.

`scale_x_continuous(limits = c(...))` filters the data. Any rows outside the range are dropped before geoms or smoothers see them. This can silently change a regression line, so it is usually the wrong choice.

```{r}
#| label: fig-axis-limits
#| fig-cap: "coord_cartesian (left) crops the view; scale limits (right) filter the data."
#| fig-alt: "Two scatter plots side by side. Both show arrival year vs total embarked with a smoother. The left plot crops the x-axis to 1700-1815 but the smoother uses all the data. The right plot filters to the same range and the smoother changes shape accordingly."
#| warning: false
#| message: false
crop <- ggplot(dat_filter, aes(arrival_year, total_embarked)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  coord_cartesian(xlim = c(1700, 1815)) +
  labs(title = "coord_cartesian (crop)")

filt <- ggplot(dat_filter, aes(arrival_year, total_embarked)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  scale_x_continuous(limits = c(1700, 1815)) +
  labs(title = "scale_x_continuous(limits=)")

crop + filt
```

## Viridis colour palettes

The default `ggplot2` colours are not colour-blind friendly. The `viridis` family of palettes is and is now included in `ggplot2`. Use `scale_fill_viridis_d()` for discrete fill (a category mapped to fill), and `scale_fill_viridis_c()` for continuous fill (a number mapped to fill); equivalents exist for `colour`.

The `option` argument picks one of five palettes:

* `"A"` magma
* `"B"` inferno
* `"C"` plasma
* `"D"` viridis (the default)
* `"E"` cividis (the most accessible)

```{r}
#| label: fig-viridis-options
#| fig-cap: "Five viridis palettes applied to the same column chart."
#| fig-alt: "Five column charts of the same data with different viridis colour options (A magma, B inferno, C plasma, D viridis, E cividis). Each uses a different colour gradient from dark to light."
make_plot <- function(opt) {
  ggplot(port_summary, aes(start_port, embarked, fill = start_port)) +
    geom_col(show.legend = FALSE) +
    scale_fill_viridis_d(option = opt) +
    labs(title = paste("option =", opt), x = NULL, y = NULL)
}

make_plot("A") + make_plot("B") + make_plot("C") +
  make_plot("D") + make_plot("E") +
  plot_layout(nrow = 2)
```

## Activities

::: {.callout-tip}
## Activity 1

Apply the `theme_classic()` theme to the column chart of port totals and edit the y-axis label to "Total embarked (people)".
:::

::: {.callout-note collapse="true"}
## Solution

```{r eval = FALSE}
ggplot(port_summary, aes(x = start_port, y = embarked, fill = start_port)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "E") +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = "Port of origin", y = "Total embarked (people)") +
  theme_classic()
```
:::

::: {.callout-tip}
## Activity 2

Replace `scale_x_continuous(limits = c(1700, 1815))` in the right-hand panel above with `coord_cartesian(xlim = c(1700, 1815))`. Confirm that the smoother shape is the same as the left panel.
:::
