# Grouped plots and overplotting {#sec-grouped-plots}

```{r setup-grouped, include=FALSE}
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(patchwork)
dat_final <- read_csv("data.csv", show_col_types = FALSE) |>
  rename(id = `Voyage ID`,
         vessel = `Vessel name`,
         start_port = `Voyage itinerary imputed port where began (ptdepimp) place`,
         purchase_place = `Voyage itinerary imputed principal place of slave purchase (mjbyptimp)`,
         total_embarked = `Total embarked`,
         total_disembarked = `Total disembarked`,
         arrival_year = `Year of arrival at port of disembarkation`,
         men = `Percent men`,
         women = `Percent women`,
         children = `Percent children`,
         died = `Slaves died during middle passage`,
         mortality = `Mortality rate`,
         outcome = `Slaves outcome label`) |>
  mutate(across(c(id, vessel, purchase_place, outcome), as.factor),
         start_port = factor(start_port,
                             levels = c("Liverpool", "London", "Bristol",
                                        sort(setdiff(unique(start_port),
                                                     c("Liverpool", "London", "Bristol"))))))

dat_filter <- dat_final |>
  filter(start_port %in% c("Liverpool", "London", "Bristol"))
```

## Intended learning outcomes

By the end of this chapter you will be able to:

* Map a categorical variable to `colour` or `fill` to display three variables on one plot
* Distinguish setting an aesthetic *inside* `aes()` from setting it as a fixed value *outside* `aes()`
* Split a plot into panels with `facet_wrap()` and `facet_grid()`
* Reduce overplotting with `alpha` and `geom_jitter()`

## Functions used

* ggplot2: `aes()`, `geom_point()`, `geom_jitter()`, `geom_smooth()`, `facet_wrap()`, `facet_grid()`, `coord_cartesian()`
* `colour` and `fill` aesthetics

## Adding a third variable with colour

So far our plots have used at most two variables. To show how the number who died on the Middle Passage varied across the period of British involvement, we plot `arrival_year` against `died` and map `start_port` to `colour`.

```{r}
#| label: fig-grouped-scatter
#| fig-cap: "Number of deaths per voyage by year of arrival, coloured by port of origin."
#| fig-alt: "Scatterplot with arrival year on the x-axis and deaths per voyage on the y-axis. Points are coloured by port of origin (Liverpool, London, Bristol). Three smoothers, one per port, run across the eighteenth century."
#| warning: false
#| message: false
ggplot(dat_filter, aes(arrival_year, died, colour = start_port)) +
  geom_point(alpha = 0.4) +
  geom_smooth() +
  scale_colour_viridis_d() +
  labs(x = "Year of arrival",
       y = "Number of deaths per voyage",
       colour = "Port of origin") +
  coord_cartesian(xlim = c(1641, 1815))
```

::: {.callout-tip}
## Inside or outside `aes()`?

* Inside `aes()`: the aesthetic varies with the data. `aes(colour = start_port)` means each port gets a different colour.
* Outside `aes()`: the aesthetic is a fixed value. `geom_point(colour = "dodgerblue")` colours all points the same regardless of port.

The same rule applies to `fill`, `shape`, `size` and `linetype`.
:::

## Facetting

`facet_wrap()` splits the plot into one panel per level of a variable. It is an alternative way to display the same three-variable relationship and is often easier to read than overlaid colours.

```{r}
#| label: fig-facet-wrap
#| fig-cap: "Same data, faceted by port of origin."
#| fig-alt: "Three scatterplots arranged vertically, one for each port of origin. Each shows arrival year against deaths per voyage with its own smoother."
#| warning: false
#| message: false
ggplot(dat_filter, aes(arrival_year, died)) +
  geom_point(aes(colour = start_port), show.legend = FALSE, alpha = 0.4) +
  geom_smooth() +
  scale_colour_viridis_d() +
  facet_wrap(~ start_port, nrow = 3) +
  labs(x = "Year of arrival",
       y = "Number of deaths per voyage",
       title = "Deaths during the Middle Passage, by port of origin") +
  coord_cartesian(xlim = c(1641, 1815))
```

The base mapping for the smoother uses no `colour` mapping, so a single smoother is drawn per facet. Inside `geom_point()` we map `colour = start_port` so the points within each facet are coloured by port even though there is only one port per panel; this is purely decorative.

* Try changing `nrow = 3` to `nrow = 1` and rendering. What happens?

::: {.callout-tip}
## Activity

Replace `facet_wrap(~ start_port, nrow = 3)` with `facet_grid(start_port ~ .)`. Read the help with `?facet_grid` to see when one is preferred over the other.
:::

## Overplotting {#sec-overplotting}

The earliest scatterplots in @sec-basic-plots had so many overlapping points that the density of the data was hidden. Two techniques help.

`alpha` makes each point semi-transparent, so dense regions appear darker.

`geom_jitter()` adds a small random offset, so points stacked on the same value are visible separately.

```{r}
#| label: fig-overplotting
#| fig-cap: "Three approaches to overplotting."
#| fig-alt: "Three scatterplots side by side. Left: opaque points overlapping heavily. Middle: transparent points revealing density. Right: jittered points spreading across the y-axis."
#| warning: false
opaque <- ggplot(dat_filter, aes(start_port, mortality)) +
  geom_point() +
  labs(title = "alpha = 1 (default)")

transparent <- ggplot(dat_filter, aes(start_port, mortality)) +
  geom_point(alpha = 0.1) +
  labs(title = "alpha = 0.1")

jittered <- ggplot(dat_filter, aes(start_port, mortality)) +
  geom_jitter(alpha = 0.2, width = 0.3, height = 0) +
  labs(title = "geom_jitter()")

opaque + transparent + jittered
```

::: {.callout-warning}
`geom_jitter()` moves points slightly. That is fine for showing density on a categorical axis, but if you jitter on a continuous axis you are misrepresenting the data. Set the `width` and `height` arguments deliberately. Here we use `height = 0` so vertical positions are unchanged.
:::
