# Reshaping data and combining plots {#sec-reshaping}

```{r setup-reshaping, include=FALSE}
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(janitor)
library(patchwork)
source("R/load_data.R")
```

## Intended learning outcomes

By the end of this chapter you will be able to:

* Reshape data from wide to long with `pivot_longer()`
* Plot grouped distributions using one mapping per categorical column rather than one column per category
* Order categorical axes deliberately with `factor()` or `fct_relevel()`
* Combine multiple plots into a single figure with the `patchwork` package

## Functions used

* tidyr: `pivot_longer()`, `drop_na()`
* forcats: `fct_relevel()`
* ggplot2: `geom_boxplot()`
* patchwork: `+`, `/`, `plot_layout()`, `plot_annotation()`

## From wide to long {#sec-wide-to-long}

The dataset has three columns recording the demographic make-up of each voyage. `men`, `women` and `children` are each a proportion of the people embarked. Most voyages in the database do not record this breakdown, so these three columns are missing for the great majority of rows. Before reshaping, we drop the voyages with no demographic data using `drop_na()`, which removes any row that is missing a value in one of the named columns.

```{r}
dat_demographic <- dat_filter |>
  drop_na(men, women, children)

head(dat_demographic[, c("start_port", "men", "women", "children")])
```

This is the *wide* form. Each demographic category has its own column. To plot them together we need the *long* form, where one column holds the category and another holds the value.

`pivot_longer()` does the conversion. We tell it which columns to gather (`cols`), the name of the new column that will hold the original column names (`names_to`) and the name of the new column that will hold the values (`values_to`).

```{r}
dat_long <- dat_demographic |>
  pivot_longer(cols = c(men, women, children),
               names_to = "demographic",
               values_to = "proportion")

head(dat_long[, c("start_port", "demographic", "proportion")])
```

Each voyage now appears three times, once per demographic category. The column is called `demographic` because it holds the categories *man*, *woman* and *child*. The values are proportions of the people embarked.

::: {.callout-note}
The demographic figures describe only the voyages that recorded them, a minority of the total. The original column names `men`, `women` and `children` are inherited from the source database. In the reshaped data, the column `demographic` holds those labels as values.
:::

## Ordering categorical axes

By default, ggplot will plot `demographic` in alphabetical order, which puts `children` first. We probably want `men`, `women`, `children` or the other way around. `fct_relevel()` from `forcats` is the easiest way.

```{r}
dat_long <- dat_long |>
  mutate(demographic = fct_relevel(demographic, "men", "women", "children"))
```

## Grouped boxplots

We can now show the distribution of the demographic proportions, broken down by port of origin, all on one plot.

```{r}
#| label: fig-grouped-box
#| fig-cap: "Demographic composition of voyages from each port."
#| fig-alt: "Grouped boxplots. The x-axis shows demographic category (men, women, children). For each category there are three boxes, one per port of origin (Liverpool, London, Bristol), distinguished by fill colour. Men account for a proportion of roughly 0.5 of those embarked on average, women about 0.3 and children under 0.2."
#| warning: false
ggplot(dat_long, aes(x = demographic, y = proportion, fill = start_port)) +
  geom_boxplot(alpha = 0.6) +
  scale_fill_viridis_d(option = "E") +
  labs(x = "Demographic category",
       y = "Proportion of those embarked",
       fill = "Port of origin")
```

Men make up the majority of those embarked on every type of voyage, but the proportion of women and children varies markedly by port.

::: {.callout-tip}
## Activity

Change the value of `alpha` and observe what happens. Try `alpha = 0.2` and `alpha = 1`.
:::

## Piping straight into ggplot

You can combine `filter()` and `ggplot()` in a single pipe rather than creating an intermediate object. The example below takes the long-format data, filters to voyages that disembarked in the Americas, and plots the result.

```{r}
#| label: fig-filtered-box
#| fig-cap: "Demographic composition of voyages that disembarked in the Americas."
#| fig-alt: "Boxplots of the proportion embarked by demographic category, restricted to voyages with outcome 'Slaves disembarked in Americas'. Men, women and children are on the x-axis; proportion on the y-axis."
#| warning: false
dat_long |>
  filter(outcome == "Slaves disembarked in Americas") |>
  ggplot(aes(x = demographic, y = proportion, fill = demographic)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_viridis_d(option = "E") +
  labs(x = NULL,
       y = "Proportion of those embarked",
       title = "Demographic composition of voyages disembarking in the Americas")
```

## Combining plots with patchwork {#sec-patchwork}

`patchwork` lets you arrange `ggplot` objects into multi-panel figures. The syntax is intuitive: `+` puts plots side by side, `/` stacks them vertically, and `plot_layout()` controls the arrangement.

First, save two plots as objects.

```{r}
#| warning: false
men_plot <- dat_long |>
  filter(demographic == "men") |>
  ggplot(aes(start_port, proportion, fill = start_port)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.6) +
  scale_fill_viridis_d(option = "E") +
  labs(x = NULL, y = "Proportion men")

children_plot <- dat_long |>
  filter(demographic == "children") |>
  ggplot(aes(start_port, proportion, fill = start_port)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.6) +
  scale_fill_viridis_d(option = "E") +
  labs(x = NULL, y = "Proportion children")
```

Then combine them.

```{r}
#| label: fig-patchwork-side
#| fig-cap: "Two boxplots side by side with patchwork."
#| fig-alt: "Two boxplots arranged horizontally. The left shows the proportion of men by port; the right shows the proportion of children by port."
#| warning: false
men_plot + children_plot
```

Stacked vertically.

```{r}
#| label: fig-patchwork-stack
#| fig-cap: "The same two boxplots stacked vertically."
#| fig-alt: "The same two boxplots arranged vertically rather than side by side."
#| warning: false
men_plot / children_plot
```

`plot_annotation()` adds a single overall title across multiple panels.

```{r}
#| label: fig-patchwork-annotated
#| fig-cap: "Combined figure with a shared title."
#| fig-alt: "Side-by-side boxplots with a shared title 'Demographic composition by port of origin'."
#| warning: false
(men_plot + children_plot) +
  plot_annotation(title = "Demographic composition by port of origin")
```

See the [`patchwork` documentation](https://patchwork.data-imaginist.com/) for more elaborate layouts, shared legends and `tag_levels` for automatic panel labels (A, B, C).

## Activities

::: {.callout-tip}
## Activity 1

Build a third plot showing the proportion of women by port, and combine all three plots (`men_plot`, `women_plot`, `children_plot`) in a single row.
:::

::: {.callout-note collapse="true"}
## Solution

```{r eval = FALSE}
women_plot <- dat_long |>
  filter(demographic == "women") |>
  ggplot(aes(start_port, proportion, fill = start_port)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.6) +
  scale_fill_viridis_d(option = "E") +
  labs(x = NULL, y = "Proportion women")

men_plot + women_plot + children_plot
```
:::

::: {.callout-tip}
## Activity 2

Reshape the data back into wide format using `pivot_wider()`. (Hint: it takes `names_from` and `values_from` arguments.)
:::
